# Predictive Modelling and Beyond

**Learning objectives:**

- How to use models’ predictions on new data
- Time series analysis
- Mixed models

## Overview of predictive modelling

In this chapter we'll talk about:

- the challenges of applying predictions to new data
- the use of time series analysis, and mixed models to anticipate the trajectory of infectious diseases and health metrics.

> By exploring the underlying patterns and analysing historical data, we estimate the disease burden and evaluate the impact of interventions on population health.


## Predicting the future

### Dengue Test Predictions for 2017-2021

Test of the Dengue’s model made with mlr3 meta-package in Chapter 8. 


```{r}
#| eval: false
new_data <- hmsidwR::infectious_diseases %>%
  arrange(year)%>%
  filter(cause_name == "Dengue",
         year>=2017,
         !location_name %in% c("Eswatini", "Lesotho")) %>%
  drop_na() %>%
  group_by(location_id) %>%
  select(-location_name, -cause_name)
```

```{r}
#| eval: false
new_pred_regr.cv_glmnet <- 
  rr1$learners[[1]]$predict_newdata(new_data,
                                    task = rr1$task)

new_pred_regr.xgboost <- 
  rr2$learners[[1]]$predict_newdata(new_data,
                                    task = rr2$task)
```

## Time series analysis

> To evaluate the evolution of the phenomenon in time

Time series data can show different characteristics, such as:

- trend
- seasonality
- cyclic patterns
- irregular fluctuations

evaluation methods:

- decomposition:
  - trend
  - seasonality
  - random fluctuations
- smoothing:
  - moving average
  - exponential smoothing
- modelling:
  - ARIMA
  - Mixed models

### SDI Time Series Analysis with Mixed Effect Models

hands-on session



## Meeting Videos {-}

### Cohort 1 {-}

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
<summary> Meeting chat log </summary>

```
LOG
```
</details>
