# Spatial Data Modelling and Visualisation

**Learning objectives:**

-   Learn how to model and visualise spatial data
-   Understand the concepts of spatial data, spatial data models, and spatial models
-   Create maps, simulate infections, and predict the spatial distribution of phenomena

## Ebola

-   Ebola is a viral disease that can cause severe illness and death
-   It is transmitted through direct contact with bodily fluids of infected individuals
-   The disease is endemic in certain regions of Africa
-   The Ebola outbreak in West Africa (2014-2016) was the largest in history


<center>

<iframe width="560" height="315"  src="https://www.youtube.com/embed/5Hj4akvDNSs?si=nV-zJWoF1HHxLsNx&amp;clip=Ugkx7c_Baq_DQKQo79khjcgnQ8_vI186GMIe&amp;clipt=EKaqAhj0gAM" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen ></iframe> 
<p><em>Video: Ebola Virus Disease outbreak in West Africa (2014–2016)</em></p>
</center>

## Spatial data and models

-   Spatial data is data that has a geographic or spatial component
-   It can be represented in various formats, including raster and vector data
-   Spatial data models are used to represent and analyse spatial data
-   Spatial models are used to predict the spatial distribution of phenomena

## Make a Map

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(sf)
library(rnaturalearth)

africa <- ne_countries(continent = "Africa", 
                       returnclass = "sf")


# Africa with colours
ggplot(data = africa) +
  geom_sf(aes(fill=name), 
          color= "white", 
          linewidth = 1,
          show.legend = F) +
  coord_sf() +
  ggthemes::theme_map() +
  labs(title = "Africa",
       subtitle = "Countries in Africa",
       caption = "Sample data from rnaturalearth") +
  # add north arrow
  ggspatial::annotation_north_arrow(which_north = "true",
                         style = ggspatial::north_arrow_fancy_orienteering()) +
  # add scale bar
  ggspatial::annotation_scale(location = "tr",
                              pad_y = unit(0.1, "cm")) 

```
## Coordinate Reference System (CRS)

```{r}
st_crs(africa)
```

### Bounding Box
```{r}
ctr_africa <- africa %>%
  filter(name == "Central African Rep.")
ctr_africa %>% st_bbox()
```


## Grid of points
```{r}
#| eval: false
df_sf <- df %>%
  st_as_sf(coords = c("longitude","latitude"),
           # or use crs = 4326
           crs = "+proj=longlat +datum=WGS84") %>%
  st_intersection(ctr_africa) %>%
  st_make_valid()

grid <- df_sf %>%
  st_bbox() %>%
  st_as_sfc() %>%
  st_make_grid(what = "centers",
               cellsize = .2,
               square = F) 
```

## Create a Raster of the Temperature
```{r}
#| eval: false

library(terra) 
raster_template <- terra::rast(nrows = 18, ncols = 36,
                          xmin = 11, xmax = 28,
                          ymin = 2, ymax = 12,
                          crs = st_crs(df_sf)$proj4string)
```


```{r}
#| eval: false
ctr_africa_raster <- rasterize(df_sf,
                               raster_template,
                               field = "temperature",
                               fun = max)
```

## Dynamics of Disease Transmission

Disease transmission can be modelled using various approaches, including:

-   Compartmental models (e.g., SIR, SEIR)
-   Agent-based models
-   Network models

### Spatial Proximity with Kriging

Kriging is a geostatistical method used to predict the value of a variable at unsampled locations based on the values at sampled locations. It is particularly useful for spatial data with spatial correlation.

```{r}
#| warning: false
#| message: false
set.seed(21082024) # set seed for reproducibility 
num_points <- 100
# simulate the presence of infection 
# 1 = "infected" and 0 = "non_infected"
longitude <- rnorm(n = num_points, mean = 20.70, sd = 1)
latitude <-  rnorm(n = num_points, mean = 6.294, sd = 1)
presence <- rbinom(100, 1, prob = 0.3)  
cases <- ifelse(presence == 1, 
                rpois(n = num_points*0.7, lambda = 10), 0)
temperature <- rnorm(n = num_points, 
                     mean = 24.7, 
                     sd = (29.2 - 20.3) / 4)

# build a dataframe  
df <- data.frame(longitude, latitude, 
                 presence, cases, temperature) 

df_sf <- df %>%
  st_as_sf(coords = c("longitude","latitude"),
           # or use crs = 4326
           crs = "+proj=longlat +datum=WGS84") %>%
  st_intersection(ctr_africa) %>%
  st_make_valid()
```

```{r}
#| warning: false
#| message: false
library(gstat)

v <-  variogram(object = cases ~  temperature, data = df_sf)

v_model <- fit.variogram(v, model = vgm("Sph"))
  
plot(v, model = v_model)
```

### Perform Kriging 

to predict the spatial distribution of the variable of interest (e.g., temperature) across a grid of points in Central African Republic.

```{r}
#| warning: false
#| message: false
library(DescTools)
set.seed(240724) # set seed for reproducibility
ctr_africa_coords <- ctr_africa %>% 
  sf::st_coordinates() %>% 
  as.data.frame() %>% 
  dplyr::select(X, Y) 
bbox <- ctr_africa %>% st_bbox()
bbox_grid <- expand_grid(x = seq(from = bbox$xmin,
                                 to = bbox$xmax,
                                 length.out = 100),
                         y = seq(from = bbox$ymin,
                                 to = bbox$ymax,
                                 length.out = 100))

ctr_africa_grid_full <- data.frame(PtInPoly(bbox_grid,
                                            ctr_africa_coords))

ctr_africa_grid <- ctr_africa_grid_full %>% 
  filter(pip == 1)

ctr_africa_grid_sf <- ctr_africa_grid %>%
  st_as_sf(coords = c("x", "y"), crs = 4326) %>%
  st_make_valid() %>%
  mutate(temperature = mean(df$temperature))
```


```{r}
#| warning: false
#| message: false
k <- gstat::gstat(formula = presence ~ temperature, 
                  data = df_sf, 
                  model = v_model)

kpred <- predict(k, newdata = ctr_africa_grid_sf)

data.frame(geo = kpred$geometry,
           var = kpred$var1.var,
           pred = kpred$var1.pred) %>% 
  head()
```


```{r}
#| warning: false
#| message: false
infected_sf <- df_sf %>%
  filter(presence == 1) %>%
  st_make_valid()
```

```{r}
#| warning: false
#| message: false
ggplot() +
  geom_sf(data = kpred, 
          aes(fill = var1.pred),
          shape=21, stroke=0.5) +
  geom_sf(data = infected_sf) +
  scale_fill_viridis_c() +
  labs(title = "Kriging Prediction in Central African Rep.") +
  theme(legend.position = "right")
```


```{r}
#| warning: false
#| message: false
ggplot() +
  geom_sf(data = kpred,           
          aes(fill = var1.var),
          shape=21, stroke=0.5) +
  geom_sf(data = infected_sf) +
  scale_fill_viridis_c() +
  labs(title = "Kriging Variance in Central African Rep.") +
  theme(legend.position = "right")
```
    
    
## Resources

[R-Spatial.org](https://r-spatial.org/)

### The {sf} package

-   The {sf} package is used for handling spatial data in R
-   It provides functions for reading, writing, and manipulating spatial data
-   It supports various spatial data formats, including shapefiles, GeoJSON, and KML

```{r}
#| eval: false
# sample important {sf} functions
library(sf)

# read shapefile
africa <- st_read("data.shp")
# write shapefile
st_write(africa, "data.shp")

# convert to sf object
africa_sf <- st_as_sf(africa)

```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>

<summary>Meeting chat log</summary>

```         
LOG
```

</details>
